---
title: "FIT5145 Assignment 3 — TikTok Virality: Data Cleaning & Feasibility Analysis"
author: "Xiaowei Xu"
date: "2026-05-22"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
```

# Workflow Overview

> **Division of labour between Python and R**
>
> | Stage | Tool | What it does |
> |-------|------|-------------|
> | 1. Data acquisition (stream 9.78 GB) | **Python** (`sample_tiktok10m_filtered.py`) | Stream HuggingFace TikTok-10M, filter by year, save **raw year-filtered selected-column** CSV. No cleaning. |
> | 2. **Data cleaning + transformation** | **R** (this file) | DAMA-aligned cleaning, type coercion, deduplication, outlier removal. |
> | 3. **Feature engineering** | **R** (this file) | Engagement rate, viral label, time features, hashtag counts. |
> | 4. **Exploratory data analysis** | **R** (this file) | Distributions, t-tests, correlation, visualisations. |
>
> Python is used **only** as a data-acquisition utility (analogous to `wget`) because R has no efficient native streamer for HuggingFace Parquet shards. All analytical work — cleaning, feature engineering, statistics, EDA — is performed in R, in line with the FIT5145 A3 specification.

This analysis uses **two complementary data sources**:

| Source | Records | Size | Role |
|--------|---------|------|------|
| Kaggle `tiktok_dataset.csv` (Huseyn, 2024) | 19,382 | 3 MB | **Primary demo** — used in the main body below |
| HuggingFace `The-data-company/TikTok-10M` | ~6.65M+ | 9.78 GB (Parquet) | **Full-scale source** — pipeline design shown in the Appendix |

**Download links:**

- Kaggle: <https://www.kaggle.com/datasets/raminhuseyn/dataset-from-tiktok/data>
- HuggingFace: <https://huggingface.co/datasets/The-data-company/TikTok-10M>

```{r libs}
library(tidyverse)
library(scales)
library(knitr)
library(corrplot)
```

---

# 1. Load & Inspect Raw Data (Kaggle dataset)

```{r load}
tt_raw <- read_csv("tiktok_dataset.csv", show_col_types = FALSE)
cat("Dimensions:", nrow(tt_raw), "rows ×", ncol(tt_raw), "columns\n")
```

```{r schema}
glimpse(tt_raw)
```

```{r missing-table}
missing_df <- tt_raw %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "missing_count") %>%
  mutate(missing_pct = round(missing_count / nrow(tt_raw) * 100, 2)) %>%
  filter(missing_count > 0)

kable(missing_df, caption = "Missing Values Before Cleaning",
      col.names = c("Column", "Missing Count", "Missing %"))
```

**Observation:** 298 rows (1.54%) are missing across 7 engagement-related columns simultaneously — suggesting these rows were not yet indexed by TikTok's API at collection time, not random missingness. They will be dropped as a block.

---

# 2. Data Cleaning Pipeline (in R)

This pipeline addresses the DAMA data-quality dimensions (Week 10 lecture). The same R functions are reused in the Appendix to clean the TikTok-10M dataset.

## 2.1 Step 1 — Drop Rows with Missing Core Metrics (Completeness)

```{r drop-missing}
tt <- tt_raw %>%
  filter(!is.na(video_view_count),
         !is.na(video_like_count),
         !is.na(video_share_count),
         !is.na(video_comment_count),
         !is.na(claim_status))

cat("After dropping missing rows:", nrow(tt), "rows retained\n")
cat("Dropped:", nrow(tt_raw) - nrow(tt), "rows\n")
```

## 2.2 Step 2 — Type Coercion & Zero-View Removal (Validity)

```{r type-fix}
tt <- tt %>%
  mutate(across(c(video_view_count, video_like_count, video_share_count,
                  video_download_count, video_comment_count,
                  video_duration_sec), as.numeric))

before <- nrow(tt)
tt <- tt %>% filter(video_view_count > 0)
cat("Zero-view rows removed:", before - nrow(tt), "\n")
```

## 2.3 Step 3 — Outlier Detection (Reasonability, 3×IQR Fence)

```{r outlier}
Q1 <- quantile(tt$video_view_count, 0.25)
Q3 <- quantile(tt$video_view_count, 0.75)
IQR_val <- Q3 - Q1
fence_lo <- Q1 - 3 * IQR_val
fence_hi <- Q3 + 3 * IQR_val

outliers_n <- sum(tt$video_view_count < fence_lo | tt$video_view_count > fence_hi)
cat("Outliers detected (3×IQR on video_view_count):", outliers_n, "\n")
cat("Lower fence:", round(fence_lo, 0), " | Upper fence:", round(fence_hi, 0), "\n")
```

None detected — engagement counts in this curated dataset are naturally bounded. The same 3×IQR step is *designed to remove* bot-inflated outliers when applied to the larger TikTok-10M raw sample (see Appendix).

## 2.4 Step 4 — Duplicate Check (Uniqueness)

```{r dedup}
dupes <- sum(duplicated(tt$video_id))
cat("Duplicate video IDs:", dupes, "\n")
tt <- tt %>% distinct(video_id, .keep_all = TRUE)
cat("Rows after deduplication:", nrow(tt), "\n")
```

---

# 3. Feature Engineering (in R)

## 3.1 Engagement Rate (Integrity)

The core feature is the **category-normalised engagement rate** — replacing raw `video_view_count` (which favours large-follower accounts):

$$\text{engagement\_rate} = \frac{\text{likes} + \text{comments} + \text{shares}}{\text{views}}$$

```{r eng-rate}
tt <- tt %>%
  mutate(
    engagement_rate = (video_like_count + video_comment_count + video_share_count) /
                       pmax(video_view_count, 1),
    log_engagement = log1p(engagement_rate)
  )
summary(tt$engagement_rate)
```

## 3.2 Binary Viral Label (Top-10% Threshold)

**Scope note.** Because the Kaggle demonstration dataset does not contain reliable category or posting-time fields, this feasibility demo uses a **global top-10%** engagement-rate threshold. The full project would compute the top-10% threshold **within each category and rolling three-month window** using TikTok-10M / API data.

```{r viral-label}
cut90 <- quantile(tt$engagement_rate, 0.90)
tt <- tt %>% mutate(viral = if_else(engagement_rate >= cut90, 1L, 0L))

cat("Top-10% cut-point:", round(cut90, 4), "\n")
cat("Viral:", sum(tt$viral), "| Non-viral:", sum(tt$viral == 0), "\n")
cat("Class ratio:", round(sum(tt$viral == 0) / sum(tt$viral), 1), ": 1\n")
```

## 3.3 Duration Bins

```{r dur-bin}
tt <- tt %>%
  mutate(dur_bin = cut(video_duration_sec,
                        breaks = c(0, 15, 30, 45, 60),
                        labels = c("0-15s", "16-30s", "31-45s", "46-60s"),
                        include.lowest = TRUE))
```

## 3.4 Cleaning Summary Table

```{r summary-table}
summary_tbl <- tibble(
  Step = c("Raw data", "Drop missing metrics", "Remove zero-view rows",
           "Outlier removal (3×IQR)", "Deduplication",
           "Feature engineering", "Final dataset"),
  Rows = c(19382, 19084, 19084, 19084, 19084, 19084, nrow(tt)),
  Columns = c(12, 12, 12, 12, 12, 15, ncol(tt)),
  `DAMA Dimension` = c("—", "Completeness", "Validity",
                       "Reasonability", "Uniqueness",
                       "Integrity", "—")
)
kable(summary_tbl, caption = "Step-by-step R Data Cleaning Pipeline")
```

---

# 4. Exploratory Data Analysis (in R)

## 4.1 Distribution of Engagement Rate

```{r dist-plot, fig.height=3.5}
ggplot(tt, aes(x = log_engagement)) +
  geom_histogram(bins = 60, fill = "#4878CF", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = log1p(cut90), colour = "red", linetype = "dashed",
             linewidth = 1.2) +
  annotate("text", x = log1p(cut90) + 0.05, y = 600,
           label = sprintf("Top 10%% cut\n(ER = %.3f)", cut90),
           colour = "red", size = 3.2, hjust = 0) +
  labs(title = "Distribution of log(1 + Engagement Rate)",
       x = "log(1 + Engagement Rate)", y = "Count") +
  theme_bw(base_size = 11)
```

## 4.2 Engagement Rate by Content Type (Claim vs Opinion)

```{r claim-plot, fig.height=3.5}
ggplot(tt, aes(x = claim_status, y = engagement_rate, fill = claim_status)) +
  geom_boxplot(outlier.alpha = 0.08, outlier.size = 0.8, width = 0.5) +
  scale_fill_manual(values = c("claim" = "#4878CF", "opinion" = "#6ACC65")) +
  scale_y_log10() +
  labs(title = "Engagement Rate by Claim Status",
       x = "Content Type", y = "Engagement Rate (log scale)") +
  theme_bw(base_size = 11) + theme(legend.position = "none")
```

```{r ttest}
t.test(engagement_rate ~ claim_status, data = tt)

tt %>%
  group_by(claim_status) %>%
  summarise(median_er = round(median(engagement_rate), 3),
            n = n()) %>%
  kable(caption = "Median engagement rate by claim status")
```

**Finding:** Claim videos show significantly higher median engagement (0.394) than opinion (0.259); t-test p < 0.001. This validates content type as a predictor.

## 4.3 Visibility Ceiling Effect (Verified Status)

```{r verified-plot, fig.height=3.5}
tt %>%
  group_by(verified_status) %>%
  summarise(median_er = median(engagement_rate),
            median_views = median(video_view_count)) %>%
  pivot_longer(c(median_er, median_views)) %>%
  ggplot(aes(x = verified_status, y = value, fill = verified_status)) +
    geom_col(alpha = 0.85, width = 0.5) +
    facet_wrap(~name, scales = "free_y",
               labeller = labeller(name = c(
                 "median_er" = "Median Engagement Rate",
                 "median_views" = "Median View Count"))) +
    scale_fill_manual(values = c("not verified" = "#4878CF", "verified" = "#D65F5F")) +
    labs(title = "Visibility Ceiling: Verified accounts get MORE views but LOWER engagement",
         x = NULL, y = NULL) +
    theme_bw(base_size = 11) + theme(legend.position = "none")
```

## 4.4 Engagement by Author Ban Status

```{r ban-status}
ban_tbl <- tt %>%
  group_by(author_ban_status) %>%
  summarise(median_er = round(median(engagement_rate), 3),
            mean_er   = round(mean(engagement_rate), 3),
            n = n()) %>%
  arrange(desc(median_er))
kable(ban_tbl, caption = "Engagement rate by author ban status")
```

**Finding:** Banned (0.396) and under-review (0.351) authors show higher median engagement than active authors (0.305) — a controversial-content effect the model must account for.

## 4.5 Engagement by Video Duration Bin

```{r dur-effect}
dur_tbl <- tt %>%
  filter(!is.na(dur_bin)) %>%
  group_by(dur_bin) %>%
  summarise(median_er = round(median(engagement_rate), 3),
            mean_er = round(mean(engagement_rate), 3),
            n = n())
kable(dur_tbl, caption = "Engagement rate by video duration bin")

dur_range_pct <- round(
  (max(dur_tbl$median_er) - min(dur_tbl$median_er)) /
   min(dur_tbl$median_er) * 100, 1)
cat("Range across duration bins (median ER): +", dur_range_pct, "%\n", sep = "")
```

**Finding:** Median engagement varies only modestly across duration bins (~+3%). Duration alone is therefore a weak predictor — supporting the report's argument for **multi-feature integration** rather than single-feature heuristics.

## 4.6 Correlation Matrix

```{r corr-plot, fig.height=4.5, fig.width=6}
num_cols <- c("video_view_count", "video_like_count", "video_share_count",
              "video_download_count", "video_comment_count")
cor_mat <- tt %>% select(all_of(num_cols)) %>% cor(use = "complete.obs")
colnames(cor_mat) <- rownames(cor_mat) <- c("Views","Likes","Shares","Downloads","Comments")
corrplot(cor_mat, method = "color", type = "lower", addCoef.col = "black",
         number.cex = 0.8, tl.cex = 0.9, cl.cex = 0.8,
         col = colorRampPalette(c("white","#4878CF"))(200),
         title = "Correlation Matrix — Engagement Metrics", mar = c(0,0,1.5,0))
```

---

# 5. Feasibility Conclusion

Findings from this R demonstration on the Kaggle dataset (n = 19,084):

1. **Tractable virality definition.** Top-10% engagement cut-point yields a clean ~10:1 class imbalance with a stable threshold.
2. **Multiple signal sources.** `claim_status` (t-test p < 0.001), `verified_status`, and `author_ban_status` all show statistically distinguishable engagement distributions, while `video_duration_sec` alone shows only ~+3% variation across bins — exactly the pattern that justifies the **multi-feature integration** novelty claim.
3. **TikTok-10M closes the remaining feature gaps.** Music/audio (`music_id`, `music_original`), temporal (`create_time`), and geospatial (`poi_*`) features unavailable in the Kaggle dataset are present in the TikTok-10M dataset — the contextual features required for the tri-modal model in the report.

---

# Appendix — R Cleaning Pipeline Designed for TikTok-10M

This appendix documents the **R cleaning pipeline** that would be applied to the year-filtered TikTok-10M data. The Python streaming step writes a raw year-filtered selected-column CSV; the R code below is intended to be run against that file. 

To produce the input for this appendix:

```bash
python probe_year_distribution.py            # see year distribution
python sample_tiktok10m_filtered.py \
       --year_start 2023 --year_end 2025 \
       --target_rows 200000
```

```{r tt10m-load, eval=FALSE}
# A.1 Load raw filtered TikTok-10M data (saved by Python streaming script)
tt10m_raw <- read_csv("tiktok10m_filtered_raw.csv", show_col_types = FALSE)
cat("Raw dimensions:", nrow(tt10m_raw), "rows ×", ncol(tt10m_raw), "columns\n")

# Year distribution (sanity check that Python filter worked)
tt10m_raw %>%
  mutate(post_year = year(as_datetime(create_time))) %>%
  count(post_year) %>%
  kable(caption = "Year distribution of raw filtered data")
```

```{r tt10m-clean, eval=FALSE}
# A.2 Apply DAMA-aligned cleaning pipeline (all in R)

# Step 1: Completeness — drop rows missing core engagement metrics
tt10m <- tt10m_raw %>%
  filter(!is.na(play_count), !is.na(digg_count),
         !is.na(comment_count), !is.na(share_count),
         !is.na(duration))
cat("After completeness filter:", nrow(tt10m), "rows\n")

# Step 2: Validity — remove zero-view records
tt10m <- tt10m %>% filter(play_count > 0)

# Step 3: Reasonability — 3×IQR outlier removal on play_count
Q1 <- quantile(tt10m$play_count, 0.25)
Q3 <- quantile(tt10m$play_count, 0.75)
IQR_val <- Q3 - Q1
tt10m <- tt10m %>%
  filter(play_count >= Q1 - 3*IQR_val,
         play_count <= Q3 + 3*IQR_val)
cat("After outlier removal:", nrow(tt10m), "rows\n")

# Step 4: Uniqueness — deduplicate on video id
tt10m <- tt10m %>% distinct(id, .keep_all = TRUE)
cat("After deduplication:", nrow(tt10m), "rows\n")

# Step 5: Integrity — boolean coercion
bool_cols <- c("is_ad", "official_item", "original_item", "duet_enabled",
               "stitch_enabled", "share_enabled", "user_verified",
               "user_tt_seller", "music_original")
tt10m <- tt10m %>%
  mutate(across(any_of(bool_cols),
                ~ case_when(. == "t" ~ TRUE, . == "f" ~ FALSE, TRUE ~ as.logical(.))))
```

```{r tt10m-engineer, eval=FALSE}
# A.3 Feature engineering — all in R

tt10m <- tt10m %>%
  mutate(
    # Engagement rate
    engagement_rate = (digg_count + comment_count + share_count) / pmax(play_count, 1),
    log_engagement  = log1p(engagement_rate),

    # Viral label (top 10%)
    viral = if_else(engagement_rate >= quantile(engagement_rate, 0.90), 1L, 0L),

    # Duration bin
    dur_bin = cut(duration, breaks = c(0, 15, 30, 45, 60, Inf),
                  labels = c("≤15s","16-30s","31-45s","46-60s",">60s"),
                  include.lowest = TRUE),

    # Time features (from unix timestamp; UTC to match the Python streamer)
    create_dt = as_datetime(create_time, tz = "UTC"),
    post_year = year(create_dt),
    post_month = month(create_dt),
    post_hour = hour(create_dt),
    post_weekday = wday(create_dt, week_start = 1),
    is_weekend = if_else(post_weekday >= 6, 1L, 0L),

    # Caption stats
    caption_length = nchar(replace_na(desc, "")),
    caption_word_count = str_count(replace_na(desc, ""), "\\S+")
  )

# Hashtag count from challenges JSON column
tt10m <- tt10m %>%
  mutate(hashtag_count = str_count(replace_na(challenges, ""), '"name"'))
```

```{r tt10m-eda, eval=FALSE}
# A.4 Exploratory analysis on year-filtered TikTok-10M sample

# Year × engagement
tt10m %>%
  group_by(post_year) %>%
  summarise(median_er = median(engagement_rate),
            n = n(),
            viral_pct = mean(viral) * 100)

# Post-hour effect (contextual feature unavailable in Kaggle dataset)
ggplot(tt10m, aes(x = factor(post_hour), y = engagement_rate)) +
  geom_boxplot(outlier.alpha = 0.05, fill = "#4878CF", alpha = 0.7) +
  scale_y_log10() +
  labs(title = "Engagement Rate by Posting Hour (UTC)",
       x = "Hour of day (0–23)", y = "Engagement rate (log)") +
  theme_bw(base_size = 11)

# Music originality effect
tt10m %>%
  group_by(music_original) %>%
  summarise(median_er = median(engagement_rate), n = n()) %>%
  kable(caption = "Engagement rate by music originality (contextual feature)")
```

The appendix is **structurally identical** to the main pipeline (§2–§4) — same DAMA dimensions, same `tidyverse` idioms, same statistical tests. The only difference is the dataset and the additional contextual features (posting time, music originality) made available by TikTok-10M.

---

# AI Usage Declaration

I used ChatGPT and Google Gemini as aids while completing this assignment, for purposes including reviewing R syntax, exploring feature-engineering patterns, and improving written clarity. All final data decisions, code, visualisations, and interpretations were checked, modified, and authored by me.
